---
title: "Final Project"
author: "Ahlam Abuawad, Lizzy Gibson & Yanelli Nunez"
date: "November 14, 2017"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(rvest)
library(httr)
library(stringr)
library(forcats)
library(tableone)
library(leaflet)

theme_set(theme_bw())
theme_update(legend.position = "bottom")
```

## Motivation

A global issue that affects the lives of individuals globally, even within developed countries like the United States, is a lack of access to clean drinking water. Contaminants such as lead are not always visible to the eye and can be deleterious to health especially to susceptible populations like that of children. Although schools should allow children safe access to drinking water, recent evaluations have found schools throughout the US to have high levels of lead beyond the set standard. The goal of this project is to investigate exposure to lead for children throughout New York City schools related to race, poverty, and state mathematics test scores. 

## Related Work

Exposure to lead in water has been an issue for a long time that has only recently received a lot of attention due to large scale exposures related to social justice issues. A recent article from [The New York Times](https://www.nytimes.com/2017/04/28/nyregion/new-york-schools-lead-water.html) describes how flushing of water outlets prior to testing temporarily decreased the amount of lead measured in NYC public schools. 

## Initial Questions

Our initial questions were whether or not exposure to lead in drinking water varies by race, poverty, disability, and ability to speak the native language in New York City public schools? And if so, by how much?

Due to information provided in both datasets, our questions evolved to determine whether or not exposures to lead in drinking water in NYC public schools related to state test scores (specifically mathematics) in addition to race and poverty. 

## Data

### Sources

All data was retrieved from the [Health Data NY Open Access Website](https://health.data.ny.gov/) API's.

* Data from [Lead Testing in School Drinking Water Sampling and Results: Most Recently Reported Beginning 2016 API](https://health.data.ny.gov/Health/Lead-Testing-in-School-Drinking-Water-Sampling-and/rkyy-fsv9).

* Data from [Lead Testing in School Drinking Water: Buildings with Lead-Free Plumbing Beginning 2016 API](https://health.data.ny.gov/Health/Lead-Testing-in-School-Drinking-Water-Buildings-wi/5hbp-c6bb).

* Data from [2015-16 Guidance Counselor Reporting - Demographic Data API](https://data.cityofnewyork.us/Education/2015-16-Guidance-Counselor-Reporting-Demographic-D/iuvu-z276).

* Data from [2013-2017 School Math Results - All API](https://data.cityofnewyork.us/Education/2013-2017-School-Math-Results-All/kha6-7i9i).

### Cleaning and Merging Data 

```{r read_lead, message = FALSE, warning = FALSE}
#read in lead data, restrict to NYC counties
lead_data <- GET("https://health.data.ny.gov/resource/ja9d-vnbi.csv", query = list(`$limit` = 5000)) %>% 
  content("parsed") %>% 
  clean_names() %>% 
  filter(county %in% c("Queens", "Kings", "New York", "Bronx", "Richmond")) %>% 
  select(school_number = school, location, any_lead_free_buildings, number_of_outlets_result_less, 
         number_of_outlets_result_greater, number_of_outlets, county, school_district)
```

```{r read_lead_free, message = FALSE, warning = FALSE}
#read in lead data, restrict to NYC counties
lead_free_data <- GET("https://health.data.ny.gov/resource/mn8r-98tx.csv", query = list(`$limit` = 5000)) %>% 
  content("parsed") %>% 
  clean_names() %>% 
  filter(county %in% c("Queens", "Kings", "New York", "Bronx", "Richmond")) %>% 
  select(school_number = school, county, school_district)
```

***THERE ARE NO SCHOOLS IN NEW YORK CITY WITH LEAD FREE PIPES***

```{r read_demo, message = FALSE, warning = FALSE}
#read in demographic characteristics, create "school_number" to merge with lead data
demo_data <- GET("https://data.cityofnewyork.us/resource/vjn2-hei2.csv", query = list(`$limit` = 2000)) %>% 
  content("parsed") %>% 
  clean_names() %>% 
  select(dbn, school_name, asian, black, english_language_learners, 
         hispanic, other_race = other, poverty, students_with_disabilities,
         white) %>% 
  mutate(asian = as.numeric(asian), asian = round(asian, 4), #convert proportions to numeric
         black = as.numeric(black), black = round(black, 4),
         english_language_learners = as.numeric(english_language_learners), 
         english_language_learners = round(english_language_learners, 4),
         hispanic = as.numeric(hispanic), hispanic = round(hispanic, 4),
         other_race = as.numeric(other_race), other_race = round(other_race, 4),
         poverty = as.numeric(poverty), poverty = round(poverty, 4),
         white = as.numeric(white), white = round(white, 4),
         students_with_disabilities = as.numeric(students_with_disabilities), 
         students_with_disabilities = round(students_with_disabilities, 4)) %>% # round proportions
  mutate(school_number = str_replace(dbn, "^..", "")) #create school number to merge with lead data
```

```{r read_scores, message = FALSE, warning = FALSE}
#read in school math scores, restrict to 2017
score_data <- GET("https://data.cityofnewyork.us/resource/stka-4ti9.csv", query = list(`$limit` = 25000)) %>% 
  content("parsed") %>% 
  clean_names() %>% 
  filter(year == 2017 & grade == "All Grades") %>% #so that testst scores are AFTER lead measurement
  select(dbn, school_name, grade, mean_scale_score)
```

* School data is restricted to average scores for all grades. This way, schools do not repeat.

```{r merge, message = FALSE, warning = FALSE}
#merge demographic characteristics with math scores
school_data <- left_join(demo_data, score_data, by = c("dbn"))

#merge demographic characteristics and math scores with lead data
all_data <- left_join(school_data, lead_data, by = "school_number") %>% 
  filter(mean_scale_score != "s") %>% 
  mutate(mean_scale_score = as.numeric(mean_scale_score),
         lead_percent = number_of_outlets_result_greater / number_of_outlets, #percent of outlets with lead
         minority = ifelse(white > 0.5, "White", "Minority")) %>% 
  select(-school_name.y, -dbn) %>% 
  mutate(location = str_replace(location, "POINT \\(", ""),
         location = str_replace(location, "\\)", "")) %>% 
  separate(location, into = c("longitude", "latitude"), sep = " ") %>% #separate lat and long
  mutate(largest = pmax(white, black, asian, hispanic, other_race)) %>% 
  mutate(race = ifelse(largest == white, "White", #create categorical race category
                      ifelse(largest == black, "Black",
                             ifelse(largest == hispanic, "Hispanic",
                                    ifelse(largest == asian, "Asian",
                                           ifelse(largest == other_race, "Other", "Missing"))))),
         latitude = as.numeric(latitude),
         longitude = as.numeric(longitude))

eda_data <- all_data %>% 
  select(school_number, english_language_learners, hispanic, other_race, poverty,
         students_with_disabilities, white, asian, black, grade, mean_scale_score,
         any_lead_free_buildings, lead_percent, minority, race)
```

### Data Dictionary

* lead_percent = proportion of all outlets with lead level > 15 ppb
* minority = "Minority" if > 50% Black, Asian, Hispanic, Other; "White" if > 50% white
* race = categorical, gives name of race variable with largest proportion

## Exploratory Analysis

Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas. 

### Table One

```{r table_one}
continuous <- c("english_language_learners", 
                "poverty","students_with_disabilities", 
                "mean_scale_score", "lead_percent")

categorical <- c("any_lead_free_buildings")

biomarkers <- c("lead_percent")

table <- CreateTableOne(vars = continuous, strata = "race", data = eda_data, factorVars = categorical)
print(table, showAllLevels = TRUE, nonnormal = biomarkers)

```

### Lead and Poverty

```{r poverty}
summary(eda_data$lead_percent)
summary(eda_data$poverty)

eda_data %>% 
  ggplot(aes(y = lead_percent, x = poverty)) + geom_hex() + 
  labs(x = "Proportion of students in poverty", y = "Proportion of pipes with lead > 15 ppb",
       title = "Poverty Status and Lead Exposure")

summary(eda_data$poverty)
summary(eda_data$lead_percent)

cor.test(eda_data$lead_percent, eda_data$poverty, method = "spearman", use = "complete.obs")
```

### Lead and Minority Status

```{r minority}
eda_data %>%
  ggplot(aes(y = lead_percent, x = race, fill = race)) + 
  geom_violin() + scale_fill_brewer(palette = "Pastel2") +
  labs(x = "Race/Ethnicity", y = "Proportion of pipes with lead > 15 ppb",
       title = "Race/Ethnicity and Lead Exposure") +
       theme(legend.position = "none")

table(eda_data$minority)
tapply(eda_data$lead_percent, eda_data$minority, mean, na.rm = TRUE)

table(eda_data$race)
tapply(eda_data$lead_percent, eda_data$race, mean, na.rm = TRUE)

wilcox.test(lead_percent ~ minority, data = eda_data)
#nonparametric TTEST

kruskal.test(lead_percent ~ as.factor(race), data = eda_data)
#non-parametric ANOVA
```

### Lead and Standardized Test Scores

```{r test}
summary(eda_data$lead_percent)
summary(eda_data$mean_scale_score)

eda_data %>% 
  ggplot(aes(y = mean_scale_score, x = lead_percent)) + 
  geom_hex() +
  labs(y = "Standardized math test scores", x = "Proportion of pipes with lead > 15 ppb",
       title = "Lead Exposure and Student Math Scores")

summary(eda_data$mean_scale_score)

cor.test(eda_data$lead_percent, eda_data$mean_scale_score, method = "spearman", use = "complete.obs")
```

#### Race and Standardized Test Scores

```{r test_race}
eda_data %>% 
  ggplot(aes(y = mean_scale_score, x = race, fill = race)) + 
  geom_violin() + scale_fill_brewer(palette = "Pastel2") +
  labs(y = "Standardized math test scores", x = "Race/Ethnicity",
       title = "Race/Ethnicity and student math scores") +
   theme(legend.position = "none")

table(eda_data$minority)
tapply(eda_data$mean_scale_score, eda_data$minority, mean, na.rm = TRUE)

table(eda_data$race)
tapply(eda_data$mean_scale_score, eda_data$race, mean, na.rm = TRUE)

t.test(mean_scale_score ~ minority, data = eda_data)

summary(aov(mean_scale_score ~ as.factor(race), data = eda_data))
```

#### Poverty and Standardized Test Scores

```{r poverty_test}
eda_data %>% 
  ggplot(aes(y = mean_scale_score, x = poverty)) + geom_hex() +
  labs(x = "Proportion of students in poverty", y = "Standardized math test scores",
       title = "Poverty Status and Student Math Scores")

cor.test(eda_data$mean_scale_score, eda_data$poverty, use = "complete.obs")

summary(eda_data$poverty)
summary(eda_data$mean_scale_score)

eda_data %>% 
  lm(mean_scale_score ~ poverty, data = .) %>% 
  broom::tidy() %>% 
  select(-std.error, -statistic) %>% 
  knitr::kable()
```

## Mapping Lead Exposures

```{r minority_schools}
all_data %>%
  filter(number_of_outlets_result_greater > 0,
         minority == "Minority") %>% 
  leaflet() %>% 
  addTiles() %>% 
  addMarkers(~longitude, ~latitude, popup = ~as.character(school_number), label = ~as.character(school_number))
```

Predominantly minorty schools with outlets that were all above 15 ppb.

```{r white_schools}
all_data %>%
  filter(number_of_outlets_result_greater > 0,
         minority == "White") %>% 
  leaflet() %>% 
  addTiles() %>% 
  addMarkers(~longitude, ~latitude, popup = ~as.character(school_number), label = ~as.character(school_number))
```


## Discussion

What were your findings? Are they what you expect? What insights into the data can you make?




