---
title: "Final Project"
author: "Ahlam Abuawad, Lizzy Gibson & Yanelli Nunez"
date: "November 14, 2017"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(rvest)
library(httr)
library(stringr)
library(forcats)
```

## Read Data

**Data from Health Data NY Open Access [Website](https://health.data.ny.gov/Health/Lead-Testing-in-School-Drinking-Water-Sampling-and/rkyy-fsv9).**

* Lead Testing in School Drinking Water Sampling and Results: Most Recently Reported Beginning 2016

```{r read_lead, message = FALSE, warning = FALSE}
#read in lead data, restrict to NYC counties
lead_data <- GET("https://health.data.ny.gov/resource/ja9d-vnbi.csv", query = list(`$limit` = 5000)) %>% 
  content("parsed") %>% 
  clean_names() %>% 
  filter(county %in% c("Queens", "Kings", "New York", "Bronx", "Richmond")) %>% 
  select(school_number = school, location, any_lead_free_buildings, number_of_outlets_result_less, 
         number_of_outlets_result_greater, number_of_outlets, county, school_district)
```

**Data from Health Data NY Open Access [Website](https://health.data.ny.gov/Health/Lead-Testing-in-School-Drinking-Water-Buildings-wi/5hbp-c6bb).**

* Lead Testing in School Drinking Water: Buildings with Lead-Free Plumbing Beginning 2016

```{r read_lead_free, message = FALSE, warning = FALSE}
#read in lead data, restrict to NYC counties
lead_free_data <- GET("https://health.data.ny.gov/resource/mn8r-98tx.csv", query = list(`$limit` = 5000)) %>% 
  content("parsed") %>% 
  clean_names() %>% 
  filter(county %in% c("Queens", "Kings", "New York", "Bronx", "Richmond")) %>% 
  select(school_number = school, county, school_district)
```

***THERE ARE NO SCHOOLS IN NEW YORK CITY WITH LEAD FREE PIPES***

**Data from NYC Open Data [Website](https://data.cityofnewyork.us/Education/2015-16-Guidance-Counselor-Reporting-Demographic-D/iuvu-z276).**

* 2015-16 Guidance Counselor Reporting - Demographic Data

```{r read_demo, message = FALSE, warning = FALSE}
#read in demographic characteristics, create "school_number" to merge with lead data
demo_data <- GET("https://data.cityofnewyork.us/resource/vjn2-hei2.csv", query = list(`$limit` = 2000)) %>% 
  content("parsed") %>% 
  clean_names() %>% 
  select(dbn, school_name, asian, black, english_language_learners, 
         hispanic, other_race = other, poverty, students_with_disabilities,
         white) %>% 
  mutate(asian = as.numeric(asian), asian = round(asian, 4), #convert proportions to numeric
         black = as.numeric(black), black = round(black, 4),
         english_language_learners = as.numeric(english_language_learners), 
         english_language_learners = round(english_language_learners, 4),
         hispanic = as.numeric(hispanic), hispanic = round(hispanic, 4),
         other_race = as.numeric(other_race), other_race = round(other_race, 4),
         poverty = as.numeric(poverty), poverty = round(poverty, 4),
         white = as.numeric(white), white = round(white, 4),
         students_with_disabilities = as.numeric(students_with_disabilities), 
         students_with_disabilities = round(students_with_disabilities, 4)) %>% # round proportions
  mutate(school_number = str_replace(dbn, "^..", "")) #create school number to merge with lead data
```

**Data from NYC Open Data [Website](https://data.cityofnewyork.us/Education/2013-2017-School-Math-Results-All/kha6-7i9i).**

* 2013-2017 School Math Results - All

```{r read_scores, message = FALSE, warning = FALSE}
#read in school math scores, restrict to 2017
score_data <- GET("https://data.cityofnewyork.us/resource/stka-4ti9.csv", query = list(`$limit` = 25000)) %>% 
  content("parsed") %>% 
  clean_names() %>% 
  filter(year == 2017 & grade == "All Grades") %>% #so that testst scores are AFTER lead measurement
  select(dbn, school_name, grade, mean_scale_score)
```

* School data is restricted to average scores for all grades. This way, schools do not repeat.

## Merge data

```{r merge, message = FALSE, warning = FALSE}
#merge demographic characteristics with math scores
school_data <- left_join(demo_data, score_data, by = c("dbn"))

#merge demographic characteristics and math scores with lead data
all_data <- left_join(school_data, lead_data, by = "school_number") %>% 
  filter(mean_scale_score != "s") %>% 
  mutate(mean_scale_score = as.numeric(mean_scale_score),
         lead_percent = number_of_outlets_result_greater / number_of_outlets, #percent of outlets with lead
         minority = ifelse(white > 0.5, "White", "Minority")) %>% 
  select(-school_name.y, -dbn) %>% 
  mutate(location = str_replace(location, "POINT \\(", ""),
         location = str_replace(location, "\\)", "")) %>% 
  separate(location, into = c("longitude", "latitude"), sep = " ") %>% #separate lat and long
  mutate(largest = pmax(white, black, asian, hispanic, other_race)) %>% 
  mutate(race = ifelse(largest == white, "White", #create categorical race category
                      ifelse(largest == black, "Black",
                             ifelse(largest == hispanic, "Hispanic",
                                    ifelse(largest == asian, "Asian",
                                           ifelse(largest == other_race, "Other", "Missing"))))),
         latitude = as.numeric(latitude),
         longitude = as.numeric(longitude))

write_csv(all_data, "./all_data.csv")
```

## Data Dictionary

* lead_percent = proportion of all outlets with lead level > 15 ppb
* minority = "Minority" if > 50% Black, Asian, Hispanic, Other; "White" if > 50% white
* race = categorical, gives name of race variable with largest proportion

```{r}
library(ggmap)

nyc_map <- get_map('new york', maptype = 'toner', zoom = 11)

ggmap(nyc_map) +
  geom_point(aes(x = longitude, y = latitude), 
                 data = all_data, alpha = .5, 
                 color = "dark red", size = 1)
```






